Efficiency

### 2018-09-26 Second.Rnw
exec summary proofed

{\huge \textsf{A response to the Government response}}
\vskip\baselineskip
{\huge \textsf{(Or lack thereof)}}
\vskip\baselineskip
{\huge \textsf{on the question of the pumpjet}}
\vskip\baselineskip
\noindent
\textbf{Aidan Morrison}
\vskip\baselineskip
\textsf{August 2018}
\vskip\baselineskip
\vskip\baselineskip

@@

\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
\color{white}
\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
\par
\noindent
\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}

@@

\section{Executive Summary}
In January 2018, I produced a detailed report\parencite{morrison2018}, commissioned by Gary Johnston from Submarines for Australia, in order to examine the question of the efficiency of a pumpjet.   Pumpjets, whilst common on nuclear submarines which can operate at high speeds (above 25kt) for sustained periods, are generally not used on diesel-electric submarine, which need to operate at very low speeds (around 10kt during transit, and under 5kt during combat operations).  Considerable emphasis had been placed on the significancw of the pumpjet proposed in successful French bid made by DCNS (now Naval Group) for Australia's Future Submarine.  I concluded that an inevitable characateristic of pumpjets - directly necessitated by the characteristics which gave them good acoustic performance at high speeds - was very low propulsive performance (efficiency) at low speeds.  This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of pumpjets.  Given the profound importance of these characteristics to all diesel-electric submarines, aswell as elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.

On June 7 2018, I appeared as a witness before the Senate Economic References Committee \parencite{SEN2018morrison}, where I defended and expanded upon my results, including in the presence of senior officials from Defence and Naval Group, who steadfastly refused to comment.  The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry Into the Future Of Naval Shipbuilding Industy in Australia on 27th of June, which they finally produced in mid September.  Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in responses to Questions on Notice in Parliament, and also in the media.  With Defence's submission finally received, this paper is in a position to comprehensively assess the totalality of the arguments put by Defence against the conclusions of my previous paper.

With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research, and instead has disgraced their technical credibility with utterly ridiculous misrepresentations of the relevant physics.  Their claim that pumpjets can be efficient across the entire speed range are instead based on the assumption that the resistance faced by fully submerged bodies is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}. This is simply untrue.  It is a bizarre, irrational claim with no basis whatsoever in physics.  It is frankly bewildering that such a claim could be made, given how easily it's falsehood can be established by even moderate research, or simple logic.  If it were true, then submarines and torpedoes would be able to elevate their speed without increased resistance, and consequently reach any speed they might choose without requiring larger motors, or more fuel, and hence attain virtually unlimited range.

The truth is that the resistance of submerged bodies in the real world scales with velocity squared.  This exponential increase in resistance explains why conventional submarines slow to extremely low speeds in order to conserve energy when dived, and has been used by other respected physicists discussing this topic \parencite{davies2017} as well as essentially all scientific literature involved hydrodynamics.  It also explains why a propulsor, which must faces one hundred times as much resistance when travelling at 30kt as it does at 3kt, behaves very differently in these regimes, and consequently has a different efficiency.  For Defence, assisted by DSTG, to claim that the relationshp between drag resistance and speed under-water is constant is an extremely shocking revelation.  In the engineering/technical community, it would be just as strange to hear that a dentist with teenage children believes in the tooth-fairy.  Not only is is their job to know otherwise, but abundant life experience would have continually proven otherwise.

@@

Such a revelation necessitates a further question regarding whether such misleading claims where made out of complete ignorance and incompetence, or whether they were deliberately intended to deceive.  As can be seen from the chronology of public comments which this report details, it is difficult to see how Defence could hold such a naive and embarrassing view after having their attention drawn to the opposing evidence multiple times, and devoted so much resources to understanding the relevant technologies.  In 2012 Senator Johnston challenged RADM Moffit directly as to why research money was even being directed towards pumpjets, given how obviously inappropriate they were for our use, but RADM Moffit insisted they would still do the work required. In 2017 Defence revealed that a 99 page report was commissioned on pumpjets as part of \$214 million in 'de-risking' research, and that over a million dollars had subsequently been spent on pumpjet research specifically.

Notwithstanding the improbability that the incompetence of Defence is so completely pervasive that all of this research failed to yeild even a basic understanding of hydrodynamics, the probability that the necessary degree of ignorance has occurred through arrangement rather than accident can be most firmly established by three simple propositions, which are made obvious through a reading of this report:

\begin{enumerate}
\item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have well-formed opinion about the performance of pumpjets on diesel-electric submarines.  They would have extensive operational performance data from their in-service SSBNS which use the pumpjet.  They would have extremely detailed modelling, and tank-test results from the pumpjet from the almost-complete Barracuda class, which has an identical hull diameter to the proposed Shortfin Barracuda. The would have extensive operational performance data from their own Rubis Class nuclear submarines, which still use propellers, as well as the results of earlier tests, confirmed to have occurred by Naval Group Officials, where a pumpjet was fitted to a diesel-electric Agosta Class submarine.  That they have adequate performance data from all the necessary reference points to assess the likely performance of the pumpjet on our own is entirely beyond dispute.

\item  \textbf{The French gave the answer to the Australians.} It would naturally be a matter of profound neglect on the part of Australian officials to fail to ascertain the necessary performance data - both acoustic and propulsive efficiency - from the French during a Competitive Evaluation Process.  Any doubt that this occurred, however, can be reliably dismissed on account of public statements made by DCNS during the CEP claiming proudly that ``The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."\parencite{costello2016}. Given RADM Sammut's repeated statements \parencite{SEN2016sammut} \parencite{SEN2017hanson} to Senate Estimates that  ``the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed" was one of the reasons that DCNS was selected as design partner, and DCNS's clear pride in  ``the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology" there can also be no doubt that after the conclusion of the CEP and the commencement of design, absolutely nothing would have prevented Australia seeking further performance details or reference points, or even assistance in interpretation in the case that the information provided in the CEP was found to be insufficient.

@@

\item \textbf{The hull and the pumpjet need to be `tuned' to each other.} This is the longest-running argument, which has been made repeatedly by RADM Sammut in sequential years \parencite{sammut2017}, \parencite{bergmann2018}, in answers to Questions on Notice \ref{QON1}, and by Chief of Maritime at DSTG \parencite{kerr2018} as well as the final response to my paper.\parencite{defence2018}.  It constitutes an attempt to deny that other nuclear submarines, including the nuclear Barracuda, serve as an adequate reference point to assess the probable performance of the pumpjet on the Shortfin Barracuda.  It requires an enormous contortion of logic order to argue that the performance of the pumpjet we will have can't be known on account of the hull.  It would be equally fair to argue that one cannot know how a pair of shoes would perform in a given circumstance without simultaneously tuning them to a pair of socks. Whilst it is true that the two have to work in close conjunction with each other and a particularly poor match might impair performance, it is necessary (and sufficient) during concept design to assume that a competent (or even ideal) match is made in order to assess the overall suitability of the system.  The implausibility of this argument is further cemented by just how much emphasis has been given to the fact that the Shortfin Barracuda will have \textit{exactly the same} diameter hull as the nuclear cousin from which it is derived \ref{SEN2017hanson}, \parencite{kerr2017}, which ought to make adjustments to the pumpjet on account of the hull entirely unwarranted.

\item \textbf{The weight of the pumpjet makes it inappropriate for smaller diesel-electric submarines, but not for larger ones.} This argument has only occurred recently, first made by former CEO of Naval Group Australia Brent Clark \parencite{clark2018}, (to which I responded\parencite{morrison2018b}), and then later explained in more detail by Chief of Maritime at DSTG Dr David Kershaw\parencite{kerr2018}.  It is an attempt to deny that any other non-nuclear submarine, including every operating class in existence (none of which pumpjets) constitute an adequate reference point for assessing the likely performance of the pumpjet on our future non-nuclear submarine.  It is demonstrably ridiculous, since if a smaller submarine could not maintain proper balance and buoyancy with a heavy pumpjet attached to its tail, then the the submarine would be doomed to either tail-stand or sink, which would amount to a catastrophic failure if ever such a submarine were to detach from its moorings. That this is not the case is amply demonstrated by the fact that both the French and Russians have succeeded in fitting pumpjets to small submarines, and such catastrophes were avoided.  The Russian test boat, B-871 `Alrosa' still operates regularly in the Black Sea.\footnote{The Black Sea is an inland sea about half the size of NSW, which shares borders only with Romania, Bulgaria, Georgia, Ukraine and Turkey. Russia operates a fleet of seven Kilo Class Submarines in the Black Sea, mostly out of Sevastopol, the port which it aquired control of through the 2014 annexation of Crimea.  The Black Sea might thus be described appropriately as a `Russian Lake', where the endurance and range penalties imposed on Alrosa by the pumpjet might be tolerated.  None of the Kilo Class Submarines produced since Alrosa have a pumpjet, and none of the 15 Russian Kilo Class submarines operating in the Northern, Pacific, or Baltic Fleets have pumpjets.}

\item \textbf{The resistance of submerged bodies is constant with speed, unlike that of ships who encounter wave-making resistance.}  This is the most recent, and the most utterly and demonstrably false claim, made both by Dr David Kershaw of DSTG in an interview with Australian Defence Magazine \parencite{kerr2018}, and in Defence's considered written submission to the Senate Economic References Committee \parencite{defence2018}.  It constitutes a clearly desperate and severely embarrassing attempt to argue the no propulsion system of any sort as tested or demonstrated on a surface vessel serves as any kind of reference point for its performance on a submarine.  It is true that ships have the \textit{additional} resistance from wave-making resistance, but for most displacement vessels operating at design speed (and certainly all submarines) pressure drag and skin friction on the submerged part of the hull tend to dominate, both of which scale with velocity squared.  Consequently, the `advance ratio' (how far forward through the water a propeller moves for a given degree of rotation) changes enormously with speed, for a submarine just as it does for a surface ship, and efficiency consequently changes along with this.

@@

As noted by Andrew Davies when he contemplated the possibility that `Naval Group is pulling the wool over the eyes of a clueless Commonwealth' with the pump jet, extraordinary claims require extraordinary evidence \parencite{davies2017}.  There can be no doubt that such extraordinary evidence of cluelessness has now been supplied by the Department of Defence themselves. But this evidence, revealing impression of cluelessness so complete, is difficult to reconcile with the adamant insistence of competence and confidence in the choice, given by multiple senior officials in the face of sustained questions backed up by substantial evidence. It is so extraordinary that it demands the question of whether the appointed officials who represent the Commonwealth are indeed `clueless' or whether officials in the Navy and Department of Defence are actually conspiring with Naval Group and the French Government to pull the wool over the eyes of a clueless Parliament, as well as the properly elected members of the executive who ultimately signed off on the purchase, and the Australian public at large.

The gravity of such a conclusion is difficult to understate, given the likely nature of the motives which would attend to such a decision.  One would presume that one would not advance into the public arena a submarine design which was known to be unviable, and lead to deep embarassment and ridicule if subjected to thorough scrutiny, unless the real motive for the design or partner choice was even less capable of withstanding public scrutiny.  The possibility that this is the case must be considered very seriously, given the difficulty of reconciling recent events with the counterfactual proposition:

After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report \ref{QON1}.  Whilst Defence avoiding answering all of them directly, one of them was particularly brazenly side-stepped.  Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.  If the answer actually was what we would all assume it should be, that should be a very easy answer to directly.  A simple `yes' would do just fine.  If the answer is `no', then we must assume that at the time of writing (presumeably before the return date of 3 April, though Defence isn't always timely) that Defence an inkling (at the very least) of what Naval Group undoubtedly knew all along - that pumpjets are entirely unsuitable for diesel-electric submarines.  In avoiding answering the question directly, Defence was not just misleading the parliament, but clearly acting to shelter Naval Group from subsequent accusations of fraudulent or misleading conduct during the CEP.

Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps they felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.  One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel, along with a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or not proceed to a build contract on completion of the design.

The state of commercial negotiatons between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, reveal that Defence was moving on exactly the opposite tack - shocking Senators with the revelation that Defence was actually moving instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming weeks, effectively wiping out the opportunity for any competitive tension at any stage of the ~\$50 billion project.

In light of these events, I find the most plausible explanation to be the worst:
\begin{itemize}
\item The propulsive performance of pumpjets do indeed decline to very low efficiencies at low speeds, making them entirely unsuitable for a diesel-electric submarine which operates almost entirely at such speeds, broadly in line with the findings of my first report.
\item Some group of individuals, including senior French officials from Naval Group as well as the Australian Government, were aware of this fact at the time of the Competitive Evaluation Process, and remains so today.
\item That group has deliberatly cultivated the impression that pump-jets would confer a substantial technological edge to Australia's Future Submarine, knowing that would not be the case, either by virtue of the pump-jets not ultimately being used, or because if they were they would confer a disadvantage.
\item That impression was necessary in order to defend of a choice which selected a design partner who's proposal was both more expensive, and slower to deliver that their competitors, for which no more defensible reason could be found.
\end{itemize}

It appears that the pumpjet idea was simply a lead-character used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.  There is almost no chance that it will be included in the detailed design.  In reality, without a nuclear reactor, there is barely any relevant technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferrable.  The extent to which the bottom of the barrel needed to be scraped was illustrated when RADM Sammut was challenged on the benefit of the nuclear design, other than pump-jets \ref{SENhanson}, and he was forced to refer to the `diameter of the hull' as being a key learning, a prospect which would be much more easily subjected to public ridicule had it been more substantially promoted.\footnote{Amongst the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.  The Collins Class was evolved from a submarine of smaller hull diameter, and this aspect of the design and development encountered no problems.  In 2006, DSTO commissioned a paper suggesting that the Collins Class replacement be designed at a wider hull, which didn't correspond to that of the Barracuda.}  In contrast, those countries whos own navies relied entirely upon diesel-electric submarines would likely boast significant advantages in technologies which are genuine game-changers for diesel-electric submarines, such as advanced composite propellers, Air-Independent Propulsion, and Lithium-Ion batteries, which we are now told won't be included in the Future Submarine \textit{because} of the pumpjet\parencite{bergmann2018}.

Australians have a right to know what the true motive of Defence was in choosing the French as a design partner.  There can be no doubt that the pumpjet played no legitimate part in it.  Any claims to the contrary are made either out of obscene incompetence, as Defence's latest response might have us believe, or more likely are made with the intent of being misleading.

@@

\end{tcolorbox}
This dialogue in the Senate Estimates received media coverage highlighting Senator Hanson's apparent lack of understanding about the difference between a submarine's power source (nuclear or diesel) and propulsion (propeller or pumpjet).  However, she succeeded in applying sustained pressure to the question of the pumpjet, and the significance that it had in the selection of DCNS as a design partner.  This elicited a committment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".

>>>>

{\huge \textsf{A reply to the Government's response}}
\vskip\baselineskip
{\huge \textsf{to the question of the pumpjet}}
\vskip\baselineskip
\noindent
\textbf{Aidan Morrison}
\vskip\baselineskip
\textsf{September 2018}
\vskip\baselineskip
\vskip\baselineskip

@@

\includegraphics[width=6cm]{tl_logo.png}\\[-1em]
\color{white}
\makebox[0pt][l]{\rule{1.3\textwidth}{1pt}}
\par
\noindent
\textbf{\textsf{Trendlock}} \textcolor{namecolor}{\textsf{Consulting}}
\vskip\baselineskip
\textsf{Prepared for Submarines for Australia}

@@

\section{Executive Summary}
In January 2018, I produced a detailed report\parencite{morrison2018}, commissioned by Gary Johnston from Submarines for Australia, in order to examine the question of the efficiency of a pumpjet.   Pumpjets, whilst common on nuclear submarines which can operate at high speeds (above 25kt) for sustained periods, are generally not used on diesel-electric submarines, which need to operate at very low speeds (around 10kt during transit, and under 5kt during combat operations).  Considerable emphasis had been placed on the significance of the pumpjet proposed in the successful French bid made by DCNS (now Naval Group) for Australia's Future Submarine.  I concluded that an inevitable characteristic of pumpjets - directly necessitated by the characteristics which gave them good acoustic performance at high speeds - was very low propulsive performance (efficiency) at low speeds.  This would mean that a diesel-electric submarine would suffer poorer range and endurance, and significantly worse dived range and endurance, as well as a worse indiscretion ratio due to the use of a pumpjet.  Given the profound importance of these characteristics to all diesel-electric submarines, as well as the elevated importance of range and endurance to the Australian requirement, it would appear that a pumpjet can be safely regarded as entirely unsuitable for Australia's Future Submarine.

On June 7 2018, I appeared as a witness before the Senate Economic References Committee \parencite{SEN2018morrison}, where I defended and expanded upon my results.  My evidence was presented in the presence of senior officials from Defence and Naval Group, who steadfastly refused to comment.  The Committee directed Defence to provide a written response to my paper prior to the conclusion of the Inquiry into the Future of Australia's Naval Shipbuilding Industry (scheduled for the 27th of June) which they finally produced in mid September.  Submarines for Australia had already commissioned me to write a subsequent report in reply to the responses that had appeared in responses to Questions on Notice in Parliament, and also in the media.  With Defence's submission finally received, this paper is in a position to comprehensively assess the totality of the arguments put by Defence against the conclusions of my previous paper.

With the receipt of the written report it has become clear that Defence has no credible basis for rejecting the findings of my research.  Their arguments are so completely incongruous with the most essential and frankly obvious physical principles, it appears that Defences has quite thoroughly disgraced their technical credibility in making this response.

Their claim that pumpjets can be efficient across the entire speed range, including patrol speed, are instead based on the assumption that the efficiency of propulsors operating on fully-submerged bodies (like submarines) is ``essentially constant over speed"\parencite{defence2018} \parencite{kerr2018}.  This is supported by the argument that efficiency relates to resistance, and that resistance faced by underwater bodies does not vary with speed, since it is only wave-making resistance (encountered by vessels on the surface) that is the speed-dependent component of resistance.

This is simply untrue.  It is a bizarre, irrational claim with no basis whatsoever in physics.  It is frankly bewildering that such a claim could be made, given how easily it's falsehood can be established by even moderate research, or simple logic.  If it were true, then submarines and torpedoes would be able to elevate their speed without increased resistance, and consequently reach any speed they might choose without requiring larger motors, or more fuel, and hence attain virtually unlimited range.

The truth is that the resistance of submerged bodies in almost all relevant real-world situations scales with velocity squared.  (Wave-making resistance is in \textit{addition} to this.) This exponential increase in resistance explains why conventional submarines slow to extremely low speeds in order to conserve energy when dived, and has been used by other respected physicists discussing this topic \parencite{davies2017} as well as essentially all scientific literature involving hydrodynamics.  It also explains why a propulsor, which must faces one hundred times as much resistance when travelling at 30kt as it does at 3kt on the same vessel, functions differently in these regimes, and consequently has a different efficiency.

For Defence, assisted by Defence Science and Technology Group (DSTG), to believe that the relationship between drag resistance and speed under-water is anywhere near constant is an extremely shocking revelation.  In the scientific community, it would be just as strange to hear that a dentist with teenage children believes in the tooth-fairy.  Not only is is their job to know better, but abundant life experience would have continually proven otherwise.

@@

Such a revelation necessitates a further question regarding whether such misleading claims where made out of complete ignorance and incompetence, or whether they were deliberately intended to deceive.  As can be seen from the chronology of public comments which this report details, it is difficult to see how Defence could hold such a naive and embarrassing view after having their attention drawn to the opposing evidence multiple times, and devoted considerable resources to understanding the relevant technologies.  In 2012 Senator Johnston challenged RADM Moffit directly as to why research money was even being directed towards pumpjets, given how obviously inappropriate they were for our use, but RADM Moffit insisted they would still do the work required to better their understanding\parencite{SE2012moffit}. In 2017 Defence revealed that a 99 page report was commissioned on pumpjets as part of \$214 million in `de-risking' research, and that over a million dollars had subsequently been spent on pumpjet research specifically.

Notwithstanding the improbability that the incompetence of Defence is so completely pervasive that all of this research failed to yield even a basic understanding of hydrodynamics, the probability that the necessary degree of ignorance has occurred through arrangement rather than accident can be most firmly established by three simple propositions:

\begin{enumerate}
\item  \textbf{The French know the answer.}  As one of only two countries in the world who have designed and produced both pumpjets and propellers, nuclear and conventional submarines, and fitted and operated each type of propulsor on both types of submarines, there can be no doubt whatsoever that Naval Group and the French Government do have a well-formed opinion about the performance of pumpjets on diesel-electric submarines.  They would have extensive operational performance data from their in-service SSBNS which use the pumpjet.  They would have extremely detailed modelling, and tank-test results from the pumpjet from the almost-complete Barracuda class, which has an identical hull diameter to the proposed Shortfin Barracuda. They would have extensive operational performance data from their own Rubis Class nuclear submarines, which still use propellers, as well as the results of earlier tests, confirmed to have occurred by Naval Group Officials\parencite{SEN2018morrison}, where a pumpjet was fitted to a diesel-electric Agosta Class submarine.  That they have adequate performance data from all the necessary reference points to assess the likely performance of the proposed pumpjet on our own submarine is entirely beyond dispute.

\item  \textbf{The French gave the answer to the Australians.} It would naturally be a matter of profound neglect on the part of Australian officials to fail to ascertain the necessary performance data - both acoustic and propulsive efficiency - from the French during a Competitive Evaluation Process.  Any doubt that this occurred, however, can be reliably dismissed on account of public statements made by DCNS during the CEP claiming proudly that ``The stealth and hydrodynamic performances of pump jet propulsion are of course classified and in Australia known only to DCNS and the Australian Government."\parencite{costello2016}. Given RADM Sammut's repeated statements \parencite{SEN2016sammut} \parencite{SEN2017hanson} to Senate Estimates that  ``the way in which technical data would be transferred to Australia and the way in which intellectual property would be managed" was one of the reasons that DCNS was selected as design partner, and DCNS's clear pride in  ``the offer from France to transfer to Australia sovereign control and use of pump-jet propulsion technology"\parencite{costello2016} there can also be no doubt that after the conclusion of the CEP and the commencement of design, absolutely nothing would have prevented Australia seeking further performance details or reference points, or even assistance in interpretation in the case that the information provided in the CEP was found to be in any way insufficient or unclear.

@@

\item \textbf{The hull and the pumpjet need to be `tuned' to each other.} This is the longest-running argument, which has been made repeatedly by RADM Sammut in sequential years\parencite{sammut2017}\parencite{bergmann2018}, in answers to Questions on Notice[\ref{QON1}], and by Chief of Maritime at DSTG \parencite{kerr2018} as well as the final response to my paper\parencite{defence2018}.  It constitutes an attempt to deny that other nuclear submarines, including the nuclear Barracuda, serve as an adequate reference point to assess the probable performance of the pumpjet on the Shortfin Barracuda.  It requires an enormous contortion of logic in order to argue that the approximate performance of the pumpjet we will have can't be known on account of the hull.  It would be equally fair to argue that one cannot know how a pair of shoes would perform in a given circumstance without simultaneously tuning them to a pair of socks. Whilst it is true that the two have to work in close conjunction with each other and a particularly poor match might impair performance, it is necessary (and sufficient) during concept design to assume that a competent (normally ideal) match is made in order to assess the overall suitability of the system.  The implausibility of this argument is further cemented by just how much emphasis has been given to the fact that the Shortfin Barracuda will have \textit{exactly the same} diameter hull as the nuclear cousin from which it is derived\parencite{SEN2017hanson}\parencite{kerr2017}, which ought to make adjustments to the pumpjet on account of the hull entirely unwarranted.

\item \textbf{The weight of the pumpjet makes it inappropriate for smaller diesel-electric submarines, but not for larger ones.} This argument has only occurred recently, first made by former CEO of Naval Group Australia Brent Clark\parencite{clark2018}, (to which I responded\parencite{morrison2018b}), and then later explained in more detail by Chief of Maritime at DSTG Dr David Kershaw\parencite{kerr2018}.  It is an attempt to deny that any other non-nuclear submarines, including every operating class in existence (none of which have adopted pumpjets) constitute an adequate reference point for assessing the likely performance of the pumpjet on our future non-nuclear submarine.  It is demonstrably ridiculous, since if a smaller submarine could not maintain proper balance and buoyancy with a heavy pumpjet attached to its tail, then the the submarine would be doomed to either tail-stand or sink, which would amount to a catastrophic failure if ever such a submarine were to detach from its moorings. That this is not the case is amply demonstrated by the fact that both the French and Russians have succeeded in fitting pumpjets to small submarines, and such catastrophes were avoided.  The Russian test boat, B-871 `Alrosa' still operates regularly in the Black Sea.\footnote{The Black Sea is an inland sea about half the size of NSW, which shares borders only with Romania, Bulgaria, Georgia, Ukraine and Turkey. Russia operates a fleet of seven Kilo Class Submarines in the Black Sea, mostly out of Sevastopol, the port which it acquired control of through the 2014 annexation of Crimea.  The Black Sea might thus be described appropriately as a `Russian Lake', where the endurance and range penalties imposed on Alrosa by the pumpjet might be tolerated.  None of the Kilo Class Submarines produced since Alrosa have a pumpjet, and none of the 15 Russian Kilo Class submarines operating in the Northern, Pacific, or Baltic Fleets have pumpjets.}

\item \textbf{The resistance of submerged bodies is constant with speed, unlike that of ships who encounter wave-making resistance.}  This is the most recent, and the most utterly and demonstrably false argument, made both by Dr David Kershaw of DSTG in an interview with Australian Defence Magazine\parencite{kerr2018}, and in Defence's considered written submission to the Senate Economic References Committee\parencite{defence2018}.  It constitutes a clearly desperate and severely embarrassing attempt to dismiss the volumous literature assessing the efficiency of propulsors as a function of `advance ratio' ($J$, a unitless metric designed specifically for the purpose of making comparisons possible across vessels encountering different levels of resistance from any source including speed) simply because such literature is often associated with ships.  It is true that ships have the \textit{additional} resistance of wave-making resistance that submarines do not, but for most displacement vessels operating at design speed (and certainly all submarines) pressure drag and skin friction on the submerged part of the hull tend to dominate, both of which scale with velocity squared.  Consequently, the `advance ratio' (how far forward through the water a propeller moves for a given degree of rotation) changes substantially with speed, for a submarine just as it does for a surface ship, and efficiency consequently changes along with this.

@@

As noted by Andrew Davies when he contemplated the possibility that ``Naval Group is pulling the wool over the eyes of a clueless Commonwealth" with the pump jet, extraordinary claims require extraordinary evidence\parencite{davies2017}.  There can be no doubt that such extraordinary evidence of cluelessness has now been supplied by the Department of Defence. There can also be no doubt that Naval Group's proposal in the Competitive Evaluation Process (CEP) to which the pumpjet was ``integral"\parencite{guillou2017} was disingenuous at best, and quite possibly fraudulent.

However, the impression of cluelessness is so severe, it is difficult to reconcile with the adamant insistence of competence and confidence in the pumpjet choice, given by multiple senior officials in the face of sustained questions backed up by substantial evidence on numerous occasions.\footnote{When Senator Carr was confirming in a public hearing with Kim Gillis (Head of Capability and Sustainment Group), that he was able to review my report and respond to my earlier evidence quickly before the close of the inquiry, Gillis replied ``The only caveat I'll put on any response we make is that the one thing we do really well is detailed engineering analysis."} It is so extraordinary that it demands the question of whether the appointed officials who represent the Commonwealth are indeed `clueless'.  The alternative proposition is that officials in the Navy and Department of Defence are complicit or co-operating with Naval Group and the French Government to pull the wool over the eyes of a clueless Parliament, including the Cabinet (who signed off on the decision), and the Australian public at large.

The gravity of reaching such a conclusion is difficult to understate, given the likely nature of the motives which would attend to such a decision.  One would presume that one would not advance into the public arena a submarine design which was known to be nonviable, and lead to deep embarrassment and ridicule if subjected to thorough scrutiny, unless the real motive for the design or partner choice was even less capable of withstanding public scrutiny.  The possibility that this is the case must be considered very seriously, given the difficulty of reconciling recent events with the counter-factual proposition:

After the initial publication of my first report, three Questions on Notice were asked by Senator Ketter, specifically referencing my report[\ref{QON1}, \ref{QON2}, \ref{QON3}].  Whilst Defence avoiding answering all of them directly, one of them was particularly brazenly side-stepped.  Asked ``Is Defence still confident that Naval Group still hold the opinion that the pump-jet is well suited to diesel-electric submarines?" Defence merely reiterated their intent is to use the pumpjets as offered, strictly avoiding mentioning anyone's opinion, either their own or Naval Group's.  If the answer actually was what we would all assume it should be, that would have been a very easy question to answer directly.  A simple `yes' would be sufficient.  If the answer was `no', then we must assume that at the time of writing (presumably before the return date of 3 April, though Defence isn't always timely) that Defence had an inkling (at the very least) of what Naval Group undoubtedly knew all along - that pumpjets are entirely unsuitable for diesel-electric submarines.  In avoiding answering the question directly, Defence was not just misleading the Parliament, but clearly acting to shelter Naval Group from subsequent questions about fraudulent or misleading conduct during the CEP.

Giving Defence the most generous possible interpretation, one might allow the possibility that Defence were just coming to grips with being completely hood-winked during the CEP by their chosen design-partner, and perhaps felt that their wrath at being deliberately deceived ought not be publicly vented, or at least not at this stage.  One might, at a stretch, allow that maintaining the opportunity for the partner to save-face was still a useful bargaining-chip that our Defence chiefs to use in order to bring a potentially treacherous contractor to heel. (Presumeably in addition to a range of contractual penalties, as well as the commercial leverage the Commonwealth would have by being able to re-introduce competition into the design process as many have suggested ought be done in any case \parencite{stanford2017}, or simply not proceed to a build contract on completion of the design.)

The state of commercial negotiations between the Commonwealth and Naval Group, as revealed in the Senate Economic References Committee \parencite{SEN2018morrison}, show that Defence was moving in exactly the opposite direction - shocking Senators with the revelation that Defence was looking instead to contract immediately with DCNS for the full construction of all 12 submarines, to the complete exclusion of the Australian Submarine Corporation, within coming months as part of the Strategic Partnering Agreement.  This would effectively wipe out the opportunity for any competitive tension at any stage of the \$79 billion five-decade project within the first two years of a five-year design phase.

In light of these events, I find the most plausible explanation to be the worst:
\begin{itemize}
\item The propulsive performance of pumpjets do indeed decline to very low efficiencies at low speeds, making them completely unsuitable for diesel-electric submarines which are forced to operate almost entirely at such speeds.  This is particularly the case for a submarine required to operate over large distances, such as that demanded by the Australian requirement.
\item Some group of individuals, including senior French officials from Naval Group as well as the Australian Government, were aware of this fact at the time of the Competitive Evaluation Process, and remains so today.
\item That group has deliberately cultivated the belief that pumpjet technology, descending directly from French experience with nuclear submarines which no other contender could claim,  would confer a substantial technological edge to Australia's Future Submarine.
\item This belief was known by that group to be false - either because pumpjets would not ultimately be used or their use would be generally disadvantageous - at the same time that is was central to justifying a choice of design partner who's proposal was substantially more expensive and slower to deliver, for which no more defensible reason could be found.
\end{itemize}

In summary, I conclude that the pumpjet was a fraudulent, rather than incomptent proposal, used to construct the fiction that France's vaunted heritage as a designer of nuclear submarines offers a substantial technical advantage to a non-nuclear submarine.  There is almost no chance that it will be included in the detailed design.  In reality, without a nuclear reactor, there is relatively little relevant nuclear-related technology or experience (other than the combat system, mandated from a 3rd party) which is usefully transferable.  The extent to which the bottom of the barrel needed to be scraped to maintain this fiction was illustrated when RADM Sammut was challenged on the benefit of the nuclear design after facing pressure on the pumpjet, and was forced to refer to the `diameter of the hull' as being a key learning[\ref{SENhanson}].  Given the relatively more accessible engineering concepts involved, such claim would be much more easily subjected to public ridicule had it been more substantially promoted, as the pumpjet was.\footnote{Among the many engineering challenges presented in designing a new submarine, ensuring the structural integrity of the hull at a new diameter is one of the most well-understood and predictable.  The Collins Class was evolved from a submarine of smaller hull diameter, and this aspect of the design and development encountered no problems.  In 2006, DSTO commissioned a paper suggesting that the Collins Class replacement be designed at a wider hull, which didn't correspond to that of the Barracuda.\parencite{joubert2006}}  In contrast, those countries whose own navies relied entirely upon diesel-electric submarines would likely boast significant advantages in technologies which are genuine game-changers for diesel-electric submarines, such as advanced composite propellers, Air-Independent Propulsion, and Lithium-Ion batteries, which we are now told won't be included in the Future Submarine \textit{because} of the pumpjet\parencite{bergmann2018}.  To that extent, it is almost certain that the pumpjet proposal will not only impose a substantial cost on Australian taxpayers (tens of billions of dollars), but also substantially degrade Australia's future military capability.

In light of these considerations, Defence should be required to explain again what their reasons were for choosing the French as a design partner, as there can now be no doubt that the pumpjet played no legitimate part in it.  Any claims to the contrary are made either out of incompetence, as Defence's latest response might have us believe, or are made with the deliberate intent of misleading.

@@

\end{tcolorbox}
This dialogue in the Senate Estimates received media coverage highlighting Senator Hanson's apparent lack of understanding about the difference between a submarine's power source (nuclear or diesel) and propulsion (propeller or pumpjet).  However, she succeeded in applying sustained pressure to the question of the pumpjet, and the significance that it had in the selection of DCNS as a design partner.  This elicited a commitment to the pumpjet that appeared hesitant in its wording, which noted that the pump jet ``remains on offer", which is what ``we intend to use".
